\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-23 paper/main.tex

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Sec.~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, \texttt{davos} will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If \texttt{False} (the default),
  \texttt{davos} will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

\item \texttt{.confirm\_install}: If \texttt{True} (default:
  \texttt{False}), \texttt{davos} will require user confirmation
  before installing a smuggled package that does not yet exist in the
  user's environment.

\item \texttt{.noninteractive}: Setting this attribute to
  \texttt{True} (default: \texttt{False}) enables non-in\-ter\-act\-ive
  mode, in which all user interactions (prompts and dialogues) are
  disabled. Note that in non-interactive mode, the
  \texttt{confirm\_install} option is set to \texttt{False}.  If
  \texttt{auto\_rerun} is \texttt{False} while in non-interactive
  mode, \texttt{davos} will raise an exception if a smuggled package
  cannot be reloaded, rather than prompting the user.

@@

Across different versions of a given package, particular modules, functions, and other objects
may be updated, removed, renamed, or otherwise altered in some manner.  In addition to changing the
behaviors of active computations, these changes can render saved objects created using one
version of a package incompatible with other versions of the
same package.  For example, the popular \texttt{pandas}~\cite{McKi10}
library used to include the \texttt{Panel} data structure for storing 3-dimensional arrays.
Since version 0.20.0, however, the \texttt{Panel} class has been deprecated, and in
version 0.25.0, it was removed entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their disk (e.g., for later reuse
or to share with other users) by serializing the \texttt{Panel} with Python's \texttt{pickle} protocol.
The \texttt{pickle} protocol is a popular, built-in method of persisting data in Python, allowing users to save, share, and load arbitrary objects.
However, in order to successfully ``unpickle'' (i.e., load and restore) a ``pickled'' object, the object's class must be defined in and importable from the same module as when it was saved.
Thus, because of the \texttt{Panel} class's removal, the user's dataset could not be read by any version of \texttt{pandas} from 0.25.0 or beyond. These
incompatibilities are also not limited solely to traditional forms of data.
For example, saved model states and other objects may reference
modules, functions, attributes, classes, or other objects that may not be identical (or even present)
across all versions of their associated package.

@@

After installing and importing \texttt{davos} (lines 1--2), we first smuggle two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} is particularly useful
for managing the installation of \texttt{pandas} in the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a pickled \texttt{Panel} object, then we must use a version of
\texttt{pandas} prior to 0.25.0 (i.e., the version in which the \texttt{Panel}
class was removed) to be able to load it in. Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, written to a CSV
file, which is compatible with newer \texttt{pandas} versions).

@@

After re-enabling these two options (line 20), we next smuggle
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 22--24). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed by line 19, though possibly as
different versions than those specified in the onion comments on lines
22 and 23. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook namespace. If they
differ, \texttt{davos} will download the requested versions in place
of the installed versions before doing so.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example5}
\end{center}
Line 24 uses an onion comment to specify that \texttt{Quail} should be
installed directly from a specific GitHub commit (\texttt{6c847a4}).
This ability to load packages directly from GitHub repositories can
enable developers to more easily use forked or modified versions of other
packages in their notebooks, even if those versions have not been
officially released.

In lines 26--29, we demonstrate another aspect of \texttt{davos}'s
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating various JavaScript widgets with Python code, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In many basic setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accomodate this multi-environment
scenario, on lines 26 and 28, we use the \texttt{pip\_executable} option to control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 30 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable more aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.

>>>>

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Sec.~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter,
  and the code that generates those objects was changed between the
  previously imported and to-be-smuggled versions.  If this attribute
  is set to \texttt{True}, \texttt{davos} will automatically restart
  the notebook kernel and rerun all code up to (and including) the
  current \texttt{smuggle} statement. If set to \texttt{False} (the default),
  \texttt{davos} will instead issue a warning, pause execution, and
  prompt the user to either restart and rerun the notebook, or
  continue running with the previously imported package version until
  the next time the kernel is restarted manually.  Note that, as of
  this writing, the \texttt{.auto\_rerun} attribute is not supported
  in Google Colaboratory notebooks.

\item \texttt{.confirm\_install}: If set to \texttt{True} (default:
  \texttt{False}), \texttt{davos} will require user confirmation
  before installing a smuggled package that does not yet exist in the
  user's environment.

\item \texttt{.noninteractive}: Setting this attribute to
  \texttt{True} (default: \texttt{False}) enables non-in\-ter\-act\-ive
  mode, in which all user interactions (prompts and dialogues) are
  disabled. Note that in non-interactive mode, the
  \texttt{confirm\_install} option is set to \texttt{False}.  If
  \texttt{auto\_rerun} is set to \texttt{False} while in non-interactive
  mode, \texttt{davos} will raise an exception if a smuggled package
  cannot be reloaded, rather than prompting the user.

@@

Across different versions of a given package, particular modules, functions, and other objects
may be updated, removed, renamed, or otherwise altered.  In addition to changing the
behaviors of active computations, these changes can render saved objects created using one
version of a package incompatible with other versions of the
same package.  For example, the popular \texttt{pandas}~\cite{McKi10}
library used to include the \texttt{Panel} data structure for storing 3-dimensional arrays.
Since version 0.20.0, however, the \texttt{Panel} class has been deprecated, and in
version 0.25.0, it was removed entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their disk (e.g., for later reuse
or to share with other users) by serializing the \texttt{Panel} with Python's \texttt{pickle} protocol.
The \texttt{pickle} protocol is a popular built-in method of persisting data in Python, allowing users to save, share, and load arbitrary objects.
However, in order to successfully ``unpickle'' (i.e., load and restore) a ``pickled'' (i.e., saved) object, the object's class must be defined in and importable from the same module as when it was saved.
Thus, because of the \texttt{Panel} class's removal, the user's dataset could not be read by any version of \texttt{pandas} from 0.25.0 or beyond. These
incompatibilities are also not limited solely to traditional forms of data.
For example, saved model states and other objects may reference
modules, functions, attributes, classes, or other objects that may not be identical (or even present)
across all versions of their associated package.

@@

After installing and importing \texttt{davos} (lines 1--2), we first \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} is particularly useful
for managing the installation of \texttt{pandas} in the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a pickled \texttt{Panel} object, then we must use a version of
\texttt{pandas} prior to 0.25.0 (i.e., the version in which the \texttt{Panel}
class was removed) to be able to load it in. Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, written to a CSV
file, which is compatible with newer \texttt{pandas} versions).

@@

After re-enabling these two options (line 20), we next \texttt{smuggle}
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 22--24). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed by line 19, though possibly as
different versions than those specified in the onion comments on lines
22 and 23. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook namespace. If they
differ, \texttt{davos} will download the requested versions in place
of the installed versions before doing so.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example5}
\end{center}
Line 24 uses an onion comment to specify that \texttt{Quail} should be
installed directly from a specific GitHub commit (\texttt{6c847a4}).
This ability to load packages directly from GitHub repositories can
enable developers to more easily use forked or modified versions of other
packages in their notebooks, even if those versions have not been
officially released.

In lines 26--29, we demonstrate another aspect of \texttt{davos}'s
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating various JavaScript widgets with Python code, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In many basic setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accomodate this multi-environment
scenario, on lines 26 and 28, we use the \texttt{pip\_executable} option to control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 31 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable more aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.
